---
title: "Elo calculations"
author: "waheeb Algabri"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(openintro)
```

#### project introduction

In this project, you’re given a text file with chess tournament results where the information has some structure. Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:
Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents
For the first player, the information would be:
Gary Hua, ON, 6.0, 1794, 1605
1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.
If you have questions about the meaning of the data or the results, please post them on the discussion forum. Data science, like chess, is a game of back and forth...
The chess rating system (invented by a Minnesota statistician named Arpad Elo) has been used in many other contexts, including assessing relative strength of employment candidates by human resource departments.
You may substitute another text file (or set of text files, or data scraped from web pages) of similar or greater complexity, and create your own assignment and solution. You may work in a small team. All of your code should be in an R markdown file (and published to rpubs.com); with your data accessible for the person running the script.


#### Loading the data

```{r}
library(stringr)
library(ggplot2)
library(tidyverse)
```

I,m going to use ./ in front of the file to make my codes more explicit and portable.

```{r}
tournment<- ("./tournamentinfo.txt")
waheeb<- readLines(tournment)
head(waheeb, 7)
```


#### Data transformation

```{r}
# remove first 4 rows that I don't need
con <- waheeb[-c(0:4)]
```

```{r}
# remove unnecessary spaces
con <- con[sapply(con, nchar) > 0]

```


```{r}
# divide odd / even rows into separate set of lines
odd <- c(seq(1, length(con), 3))
odd_a <- con[odd]

even <- c(seq(2, length(con), 3))
even_a <- con[even]
```

I will use regex to extract the only required information.

```{r}
# name
name <- str_extract(odd_a, "\\s+([[:alpha:]- ]+)\\b\\s*\\|")
name <- gsub(name, pattern = "|", replacement = "", fixed = T)
# strip the space
name <- trimws(name)

# state
state <- str_extract(even_a, "[[:alpha:]]{2}")

# total_points
total_points <- str_extract(odd_a, "[[:digit:]]+\\.[[:digit:]]")
total_points <- as.numeric(as.character(total_points))

# pre_rating
pre_rating <- str_extract(even_a, ".\\: \\s?[[:digit:]]{3,4}")
pre_rating <- gsub(pre_rating, pattern = "R: ", replacement = "", fixed = T)
pre_rating <- as.numeric(as.character(pre_rating))

# opponent_number to extract opponents pair number per player
opponent_number <- str_extract_all(odd_a, "[[:digit:]]{1,2}\\|")
opponent_number <- str_extract_all(opponent_number, "[[:digit:]]{1,2}")
opponent_number <- lapply(opponent_number, as.numeric)
```


calculate Average Pre Chess Rating of Opponents and store that in a list.

```{r}
opp_avg_rating <- list()
for (i in 1:length(opponent_number)){
  opp_avg_rating[i] <- round(mean(pre_rating[unlist(opponent_number[i])]),2)
}
opp_avg_rating <- lapply(opp_avg_rating, as.numeric)
opp_avg_rating <- data.frame(unlist(opp_avg_rating))
head(opp_avg_rating)
```


create data frame 

```{r}
df <- cbind.data.frame(name, state, total_points, pre_rating, opp_avg_rating)
colnames(df) <- c("Name", "State", "Total_points", "Pre_rating", "Avg_pre_chess_rating_of_opponents")
head(df)
```

#### Visualization

```{r}
ggplot(data = df, aes(x = Pre_rating, y = Total_points)) +
  geom_point(size = 4, color = "blue") +
  ggtitle("Pre-chess rating vs Total points earned") +
  xlab("Pre-chess rating") +
  ylab("Total points earned")

```


```{r}
df_state_points <- df %>% group_by(State) %>% 
  summarize(Total_points = sum(Total_points))

ggplot(data = df_state_points, aes(x = "", y = Total_points, fill = State)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  ggtitle("Distribution of Total points earned by players from different states") +
  labs(fill = "State") +
  scale_fill_brewer(palette = "Set1")

```

```{r}
write.csv(df, "chess_ratings.csv")
```



***

Based on difference in ratings between the chess players and each of their opponents in our Project 1 tournament, calculate each player’s expected score (e.g. 4.3) and the difference from their actual score (e.g 4.0). List the five players who most overperformed relative to their expected score, and the five players that most underperformed relative to their expected score.
You’ll find some small differences in different implementation of ELO formulas.You may use any reasonably-sourced formula, but please cite your source.



```{r}
# Import the data from CSV file
chess_data <- read.csv("chess_ratings.csv")
```



```{r}

# Calculate the performance ratings for each player
chess_data$perf_rating <- chess_data$Pre_rating + (chess_data$Total_points - 3) * 100

# Calculate the average performance rating of opponents for each player
opponents <- subset(chess_data, select=c("Name", "Avg_pre_chess_rating_of_opponents"))
colnames(opponents) <- c("Opponent", "Avg_rating")
chess_data$Opponent_rating <- merge(chess_data, opponents, by.x="Name", by.y="Opponent")$Avg_rating
chess_data$avg_perf_opponents <- round(mean(chess_data$Opponent_rating), 2)

# Print the output
head(chess_data)

```



```{r}
# calculate expected score and difference from actual score for each player
chess_data$expected_score <- 1 / (1 + 10^((chess_data$Opponent_rating - chess_data$perf_rating)/400))
chess_data$score_diff <- chess_data$Total_points - chess_data$expected_score
```



```{r}
# sort by overperformers
overperformers <- head(chess_data[order(chess_data$score_diff, decreasing = TRUE), ], 5)

```


```{r}
# sort by underperformers
underperformers <- head(chess_data[order(chess_data$score_diff), ], 5)

```


```{r}
# print results
cat("Five players who most overperformed relative to their expected score:\n")
```

```{r}
print(overperformers)
```



```{r}

cat("\nFive players who most underperformed relative to their expected score:\n")

```

```{r}
print(underperformers)
```



* You may use any reasonably-sourced formula, but please cite your source.

This is the website which I got some useful steps for the Elo rating system

[ FiveThirtyEight ]( https://fivethirtyeight.com/features/how-we-calculate-nba-elo-ratings/)


